---
title: "Plotathon"
author: "Isaac Quintanilla Salinas"
format: 
  revealjs:
    width: 1200
    scrollable: true
    theme: [default, styles.scss]
    controls-tutorial: true
    incremental: false 
    touch: false
    controls: true
    pointer:
      pointerSize: 32
    slide-number: true
    sc-sb-title: h1
    footer: inqs.info/p/plotathon/f25
knitr:
  opts_chunk: 
    echo: true
    eval: false
    message: false
    warnings: false
    comment: "#>" 

revealjs-plugins:
  - pointer
  - verticator
  
filters: 
  - reveal-header
  - code-fullscreen
  - reveal-auto-agenda
  
auto-agenda:
  bullets: numbered
  clickable: true
  heading: Outline

editor: source
---

# Installation

## Installing R

<https://cloud.r-project.org/>

## Installing RStudio

<https://posit.co/download/rstudio-desktop/>

## Installing RTools (Windows)

<https://cran.r-project.org/bin/windows/Rtools/>

## Posit Cloud

<https://posit.cloud/>

# R Packages

## R Packages


:::: panel-tabset

### Tidyverse

```{r}
install.packages("tidyverse")
```

### Plotly

```{r}
install.packages("plotly")
```

### ggthemes

```{r}
install.packages("ggthemes")
```

### Esquisse

```{r}
install.packages("esquisse")
```

### All/Other

```r
install.packages(c("tidyvere", "plotly", 
                   "ggthemes", "esquisse",
                   "ggwordcloud", "ggbump",
                   "ggpie", "tidytext", "tidytuesdayR"))
```

::::

## Load Packages

```{r}
library(tidyverse)
library(plotly)
library(ggthemes)
library(esquisse)
```

# R Projects

## R Projects

R Projects are designed to contain all your data, scripts, and output in one location. This ensures that things are compatible on other computers.

## Directories (Folders)

-   data

-   output

-   scripts

# RStudio Layout

## RStudio Layout

-   Scripts

-   Console

-   Environment

-   Files

-   Everything else

# Scripts

## Adding Script

-   Use the top right corner

# Loading Data

## Downloading Data

```{r}
#| echo: true
#| eval: false
u <- "https://www.inqs.info/p/plotathon/owenWilsonWows.csv"
basename(u)
download.file(u,
              file.path(getwd(), "data", basename(u)))

```

## Loading Data

```{r}
#| eval: false
mds <- read_csv("data/owenWilsonWows.csv")
```

# Plotting

## Base Plot

```{r}
mds |> ggplot(aes(x = total_wows_in_movie)) 
```

## Histogram

```{r}
mds |> ggplot(aes(x = total_wows_in_movie)) +
  geom_histogram()
```

## Box Plot

```{r}
mds |> ggplot(aes(x = total_wows_in_movie)) +
  geom_boxplot()
```

## Density Plot

```{r}
mds |> ggplot(aes(x = total_wows_in_movie)) +
  geom_density()
```

## Box Plot By Category

```{r}
mds |> ggplot(aes(x = total_wows_in_movie, y = era)) +
  geom_boxplot()
```

## Density Plot By Category

```{r}
mds |> ggplot(aes(x = total_wows_in_movie, color = era)) +
  geom_density()
```

## Scatter Plot

```{r}
mds |> ggplot(aes(x = durationInSeconds, 
                  y = total_wows_in_movie)) +
  geom_point()
```

## Scatter Plot by Group

```{r}
mds |> ggplot(aes(x = durationInSeconds, 
                  y = total_wows_in_movie, color = era)) +
  geom_point()
```

## Add Regression Line

```{r}
mds |> ggplot(aes(x = durationInSeconds, 
                  y = total_wows_in_movie)) +
  geom_point() +
  geom_smooth(method = "lm", se = F)
```

## Smooth Line

```{r}
mds |> ggplot(aes(x = durationInSeconds, 
                  y = total_wows_in_movie)) +
  geom_point() +
  geom_smooth(se = F)
```

## Regression Lines by Group

```{r}
mds |> ggplot(aes(x = durationInSeconds, 
                  y = total_wows_in_movie,
                  color = era)) +
  geom_point() +
  geom_smooth(method = "lm", se = F)
```

# Esquisse

# Generative Artificial Intelligence

## AI Misuse

::: incremental

- **Never upload any data to a company**

- **AI is known to make things up**

- **Do not implement a method that you know nothing about**

- **Consider who you are harming when you use AI**

:::


## AI use in R

::: incremental

- Ask AI to give you example code

- Verify with reputable resources if the code is correct

  - Read documentation

  - Identify real-life use

- Change the code to your needs


:::

# Resources

## Resources

[R Graphics Cookbook](https://r-graphics.org/)

[R Graph Gallery](https://r-graph-gallery.com/)

[R Charts](https://r-charts.com/)

[ggplot2](https://ggplot2.tidyverse.org/)

[Graphical Data Analysis in R](http://www.gradaanwr.net/)

[Modern Data Visualization](https://rkabacoff.github.io/datavis/)

## Tidy Tuesday

Tidy Tuesday is a weekly data visualization challenge from the [Data Science Learning Community](https://github.com/rfordatascience/tidytuesday).

Search #tidytuesday on social media such as Bluesky or [Fosstodon](https://fosstodon.org/tags/tidytuesday)

Also check out my attempts at tidy tuesday: [inqs.info/ttw](https://www.inqs.info/ttw)
