---
title: "Overview of Joint Longitudinal-Survival Models: Modeling the Association Between Dependent Outcomes"
author:
  - name: '**Isaac Quintanilla Salinas**'
    affil: 1
    main: true
    email: isaac.quintanillasalinas@ucr.edu
  - name: "Esra Kürüm"
    affil: 1
    main: true
    email: 'esra.kurum@ucr.edu'
affiliation:
  - num: 1
    address: Department of Statistics, University of California - Riverside
column_numbers: 4
output: 
  posterdown::posterdown_html:
    self_contained: false
    pandoc_args: --mathjax
    number_sections: false
---

```{r setup, include=FALSE, cache=TRUE}
knitr::opts_chunk$set(
  cache = FALSE,
  echo = FALSE,
  message = FALSE, 
  warning = FALSE,
  fig.show = TRUE)

library(survival)
library(nlme)
library(JMbayes)
library(DT)

me <- lme(CD4 ~ obstime + obstime:drug,
          random = ~ obstime | patient, 
          data = aids)


ph <- coxph(Surv(Time, death) ~ drug,
            data = aids.id,
            x=TRUE)
jm.1 <- JMbayes::jointModelBayes(me, ph, timeVar = "obstime")
fp <- aids[aids$patient==325,]
surv_fp <- JMbayes::survfitJM(jm.1, fp, idVar = "patient")

```


```{r rpacks, include=FALSE}
library(survival)
library(nlme)
library(JMbayes)
library(DT)
```


```{r, load_refs, include=FALSE, cache=FALSE}
library(RefManageR)
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           cite.style = "authoryear",
           style = "markdown",
           hyperlink = FALSE,
           dashed = FALSE,
           no.print.fields=c("doi","issn"))
mybib <- ReadBib("./MyBib.bib", check = FALSE)
```


```{css, echo = FALSE}
.scrollable-bib {
border: none;
padding: 5px;
font: 24px/36px sans-serif;
width: 1000px;
height: 600px;
overflow-y: scroll;
} 

/* Style the tab */
.tab {
  overflow: hidden;
  border: 1px solid #ccc;
  background-color: #f1f1f1;

}

/* Style the buttons that are used to open the tab content */
.tab button {
  background-color: inherit;
  float: left;
  border: none;
  outline: none;
  cursor: pointer;
  padding: 14px 16px;
  transition: 0.3s;
  font: 24px/36px sans-serif;
}

/* Change background color of buttons on hover */
.tab button:hover {
  background-color: #ddd;
}

/* Create an active/current tablink class */
.tab button.active {
  background-color: #ccc;
}

/* Style the tab content */
.tabcontent {
  display: none;
  padding: 6px 12px;
  border: 1px solid #ccc;
  border-top: none;
}

```

# Introduction

# Joint Longitudinal-Survival Models

### Joint Density Function

<div class="tab">
  <button class="tablinks" onclick="openCity(event, 'Rome')">Joint Density Function</button>
  <button class="tablinks" onclick="openCity(event, 'Cape Town')">Longitudinal</button>
  <button class="tablinks" onclick="openCity(event, 'Buenos Aires')">Survival</button>
  <button class="tablinks" onclick="openCity(event, 'New York')">Likelihood</button>
</div>

<div id="Rome" class="tabcontent">


$$
\begin{eqnarray}
\tiny P(T_i, \delta_i, \boldsymbol Y_i; \boldsymbol \theta) &\tiny =& \tiny\int   P(T_{i},\delta_{i}|b_{i};\boldsymbol \theta)P(\boldsymbol Y_{i}|b_{i};\boldsymbol \theta)P(b_{i};\boldsymbol \theta) db_i\\
&\tiny = & \tiny \int P(T_{i},\delta_{i}|b_{i};\boldsymbol \theta) \prod^{n_i}_j P(\boldsymbol Y_{ij}|b_{i};\boldsymbol \theta)P(b_{i};\boldsymbol \theta) db_i
\end{eqnarray}
$$
</div>

<div id="Rome" class="tabcontent">


$$
\begin{eqnarray}
\tiny P(T_i, \delta_i, \boldsymbol Y_i; \boldsymbol \theta) &\tiny =& \tiny\int   P(T_{i},\delta_{i}|b_{i};\boldsymbol \theta)P(\boldsymbol Y_{i}|b_{i};\boldsymbol \theta)P(b_{i};\boldsymbol \theta) db_i\\
&\tiny = & \tiny \int P(T_{i},\delta_{i}|b_{i};\boldsymbol \theta) \prod^{n_i}_j P(\boldsymbol Y_{ij}|b_{i};\boldsymbol \theta)P(b_{i};\boldsymbol \theta) db_i
\end{eqnarray}
$$
</div>

<div id="Cape Town" class="tabcontent">


$$
\begin{aligned}\tiny  P(\boldsymbol Y_i|b_i;\boldsymbol \theta)P(b_i;\boldsymbol \theta) = & \tiny \prod_{j=1}^{n_i}\frac{1}{\sqrt{2\pi \sigma²}}\exp\left[-\frac{1}{2\sigma²}\left\{Y_{ij}-(\boldsymbol{X}_{ij}^\mathrm T\boldsymbol \beta + \boldsymbol Z_{ij}^\mathrm Tb_i)\right\}^2\right] \\
\tiny \times & \tiny (2\pi)^{-q/2}|G|^{-1/2}\exp\left(-\frac{1}{2}b_i
^\mathrm T G ^{-1}b_i\right)\end{aligned}
$$
</div>

<div id="Buenos Aires" class="tabcontent">


$$
\begin{eqnarray}
\tiny P(T_{i},\delta_{i} |b_{i};\boldsymbol \theta)&\tiny =& \tiny \lambda_{i}(T_{i})^{\delta_{i}} S_{i}(T_{i})\notag\\
&\tiny =&\tiny \left[\lambda_0(T_{i})\exp\left\{\boldsymbol X_{i1}^\mathrm T\boldsymbol \gamma+\alpha m_i(T_i)\right\}\right]^{\delta_{i}}\notag\\
& & \tiny \times \exp\left[-\int^{T_{i}}_0\lambda_0(s)\exp\{\boldsymbol X_{i1}^\mathrm T\boldsymbol \gamma+\alpha m_i(s)\}ds\right]\\ \end{eqnarray}
$$
</div>


<div id="New York" class="tabcontent">


$$
\begin{eqnarray}
\tiny \ell(u,\boldsymbol \theta) &\tiny  = & \tiny \sum^n_{i=1}\ell_i(b_i,\boldsymbol \theta)=\sum^n_{i=1}\log P(T_{i},\delta_{i},\boldsymbol Y_{i},b_{i};\boldsymbol \theta) \\
\tiny \log P(T_{i},\delta_{i},\boldsymbol Y_{i},b_{i};\boldsymbol \theta)& \tiny = & \tiny \int \delta_{i}\left[\log\{\lambda_0(T_{i})\}+\boldsymbol X_{i1}^\mathrm T\boldsymbol \gamma+\alpha m_i(t)\right]\\
& & \tiny - \int^{T_{i}}_0\lambda_0(s)\exp\{\boldsymbol X_{i1}^\mathrm T\boldsymbol \gamma+\alpha m_i(t)\}ds \\
& & \tiny + \sum^{n_i}_{j=1}\left[- \frac{1}{2} \log(2\pi\sigma²) - \frac{1}{2\sigma²} \left\{Y_{ij}-(\boldsymbol{X}_{ij}^\mathrm T\boldsymbol \beta + \boldsymbol Z_{ij}^\mathrm Tb_i)\right\}^2\right] \\
& & \tiny - \frac{q}{2}\log(2\pi\sigma^2) - \frac{1}{2} \log(|G|)-\frac{1}{2}b_i
^\mathrm T G ^{-1}b_i db_i
\end{eqnarray}
$$
</div>


#### Submodels

### Model Specification

### Likelihood Formulation

### Estimation

# Dynamic Predictions

# Applications in R

<div class="tab">
  <button class="tablinks" onclick="openCity(event, 'London')">Longitudinal</button>
  <button class="tablinks" onclick="openCity(event, 'Paris')">Survival</button>
  <button class="tablinks" onclick="openCity(event, 'Tokyo')">Joint Model</button>
  <button class="tablinks" onclick="openCity(event, 'Los Angeles')">Plotting</button>
</div>

<div id="London" class="tabcontent">
```{r, eval=FALSE, echo = TRUE}
me <- lme(CD4 ~ obstime + obstime:drug,
          random = ~ obstime | patient, 
          data = aids)
```

</div>

<div id="Paris" class="tabcontent">
```{r, echo=T, eval=FALSE}
ph <- coxph(Surv(Time, death) ~ drug,
            data = aids.id,
            x=TRUE)
```
</div>

<div id="Tokyo" class="tabcontent">
```{r, echo=TRUE, eval=FALSE}
jm.1<- jointModelBayes(me, ph, 
                       timeVar = "obstime")
```

</div>

<div id="Los Angeles" class="tabcontent">
```{r, echo=TRUE, eval=FALSE}
JMbayes::plot.survfit.JMbayes(surv_fp,
              estimator = "median", 
              conf.int=TRUE, include.y=T)
```
</div>

<script>
function openCity(evt, cityName) {
  var i, tabcontent, tablinks;
  tabcontent = document.getElementsByClassName("tabcontent");
  for (i = 0; i < tabcontent.length; i++) {
    tabcontent[i].style.display = "none";
  }
  tablinks = document.getElementsByClassName("tablinks");
  for (i = 0; i < tablinks.length; i++) {
    tablinks[i].className = tablinks[i].className.replace(" active", "");
  }
  document.getElementById(cityName).style.display = "block";
  evt.currentTarget.className += " active";
}
</script>
   
# Conclusion

# Acknowledgements

# References

<div class="scrollable-bib">
```{r refs, echo=FALSE, results="asis"}
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           cite.style = "authoryear",
           style = "markdown",
           hyperlink = FALSE,
           dashed = FALSE,
           no.print.fields=c("issn","isbn"))
NoCite(mybib)
PrintBibliography(mybib)
```
</div>