---
title: "Data Manipulation"
author: "Isaac Quintanilla"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: 16:9
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
library(tidyverse)
select<-dplyr::select
```

## Presentation

This presentation is available at:

[gitlab.com/iquintanilla/hiss](https://gitlab.com/iquintanilla/hiss)

---

## dplyr

- Known as the Grammar of Data Manipulation

- dplyr.tidyverse.org

---

## dplyr Functions

- `mutate()` adds new variables

- `select()` selects variables

- `filter()` filters data

- `if_else()` conditional function that returns 2 values

- `group_by()` a dataset is grouped by factors

- `summarise()` provides summaries of data

---

## tidyr

- Used to create tidy data

- tidyr.tidyverse.org

---

## tidyr Functions

- `pivot_longer()` (formerly `gather()`) transforms the data from wide to long

- `pivot_wider()` (formerly `spread()`) transforms the data from long to wide

- `separate()` seperates a one variable to multiple variables

- `unite()` merge multiple variable to one variable

---

## Pipe Operator `%>%`

- The pipe operator is the real power of tidyverse.

- It takes the output of a function and uses it as input for another function.

- Tidyverse works best when data frames (tibbles) are used a inputs.

---

## `mutate()`

- Adds a new variable to a data frame 

- Example:

```{r}
mtcars %>% mutate(log_mpg=log(mpg)) %>% head()
```

---

## `mutate()`

- Each argument is a new variable added 

- Example:

```{r}
mtcars %>% mutate(log_mpg=log(mpg),log_hp=log(hp)) %>% head()
```


---

## `select()`

-This selects the variables to keep in the data frame

-Example:

```{r}
mtcars %>% mutate(log_mpg=log(mpg),log_hp=log(hp)) %>% 
  select(mpg,log_mpg,hp,log_hp) %>% head()
```

---

## `filter()`

- Selects observations that satisfy a condition

- Example:

```{r}
mtcars %>% mutate(log_mpg=log(mpg),log_hp=log(hp)) %>% 
  select(mpg,log_mpg,hp,log_hp) %>% 
  filter(log_hp<5) %>% head()
```

---

## `if_else()`

- A function that provides T (1) if the condition is met and F (0) otherwise

- Example:

```{r}
mtcars %>% mutate(log_mpg=log(mpg),log_hp=log(hp)) %>% 
  select(mpg,log_mpg,hp,log_hp) %>% 
  filter(log_hp<5) %>% 
  mutate(hilhp=if_else(log_hp>mean(log_hp),1,0)) %>% head()
```

---

## `group_by()`

- This groups the data frame

- Example:

```{r}
mtcars %>% mutate(log_mpg=log(mpg),log_hp=log(hp)) %>% 
  select(mpg,log_mpg,hp,log_hp) %>% 
  filter(log_hp<5) %>% 
  mutate(hilhp=if_else(log_hp>mean(log_hp),1,0)) %>% 
  group_by(hilhp) %>% head()
```

---

## `summarise()`

- Creates summary statistics for variables

```{r}
mtcars %>% mutate(log_mpg=log(mpg),log_hp=log(hp)) %>% 
  select(mpg,log_mpg,hp,log_hp) %>% 
  filter(log_hp<5) %>% 
  mutate(hilhp=if_else(log_hp>mean(log_hp),1,0)) %>% 
  group_by(hilhp) %>% 
  summarise(mean_mpg=mean(mpg),mean_lmpg=mean(log_mpg),
            sd_mpg=sd(mpg),sd_lmpg=sd(log_mpg)) %>% head()
```


---

## Long Data to Wide Data Example

For the rest of this presentation we work on converting data from wide to long using the functions in the tidyr package. For many statistical analysis, long data is necessarry.

---

## Load Data


Use the `read_csv()` to read 'data_3_4.csv' into an object called 'data1';

```{r,message=F}
data1<-read_csv(file="data_3_4.csv")
```


---

## Wide Data


```{r}
names(data1)
data1[1,c(1:7)]
```

---

## Long Data

```{r,include=FALSE}
data1_long <- data1 %>% pivot_longer(`v1/mean`:`v4/median`,"measurement","value") %>% 
          separate(measurement,c("time","stat"),sep="/") %>% 
          pivot_wider(names_from = stat,values_from = value)
```

```{r}
data1_long[c(1:4),c(1:5)]
```


---

## `pivot_longer()`

- The `pivot_longer()` function grabs the variables that repeated in an observation places them in one variable:


```{r}
data1 %>% 
  pivot_longer(cols=`v1/mean`:`v4/median`,names_to = "measurement",values_to = "value") %>% 
  head()
```

---

## `separate()`

- The `separate()` function will seperate a variable to multiple variables:

```{r}
data1 %>% 
  pivot_longer(cols=`v1/mean`:`v4/median`,names_to = "measurement",values_to = "value") %>% 
  separate(col=measurement,into=c("time","stat"),sep="/") %>% 
  head()
```

---

## `pivot_wider()` 

- The `pivot_wider()` function then converts long data to wide data.

```{r}
data1 %>% 
  pivot_longer(`v1/mean`:`v4/median`,"measurement","value") %>% 
  separate(measurement,c("time","stat"),sep="/") %>% 
  pivot_wider(names_from = stat,values_from = value) %>% 
  head()      
```


---


## Base R vs Tidyverse

- They both achieve the same outcomes

- Base R has more capabilities

- Tidyverse easier to read and memorize

- Example:
  - Create a table to showing the mean of `mpg` from the `mtcars` dataset grouped by the number of `cyl` and `gear.`


---

## Base R vs Tidyverse

Use the `tapply()` function in Base R:

```{r, echo=TRUE}
tapply(mtcars$mpg, list(mtcars$cyl,mtcars$gear), mean)
```

---

## Base R vs Tidyverse

Use a selection `group_by()` and `summarise()` in Tidyverse:

```{r, echo=TRUE}
mtcars %>%
  group_by(cyl,gear) %>% 
  summarise(ave=mean(mpg))
```

---

## Base R vs Tidyverse

Use the `pivot_wider()` to to force one variable to multiple variables:

```{r, echo=TRUE}
mtcars %>%
  group_by(cyl,gear) %>% 
  summarise(ave=mean(mpg)) %>% 
  pivot_wider(names_from = gear,values_from = ave) %>% 
  rename("cyl/gear"=cyl)
```


---

## Tidyverse Advantage

What if we want to know more summary statistics, use the `unite()` function to combine multiple variables to one variable:

```{r, echo=TRUE}
mtcars %>%
  group_by(cyl,gear) %>% 
  summarise(ave=mean(mpg),sd=round(sd(mpg),2)) %>% 
  unite(ave,sd,col="mean_sd",sep="/")

```


---

```{r, echo=TRUE}
mtcars %>%
  group_by(cyl,gear) %>% 
  summarise(ave=mean(mpg),sd=round(sd(mpg),2)) %>% 
  unite(ave,sd,col="mean_sd",sep="/") %>% 
  pivot_wider(names_from = "gear",values_from = "mean_sd")

```


---

## Thank You!

---

## Resources

- https://www.tidyverse.org/

- https://bookdown.org/yihui/rmarkdown/

- https://rstudio.com/

- https://slides.yihui.org/xaringan/#1

- https://remarkjs.com/#1
