---
title: "Plotathon"
author: "Isaac Quintanilla Salinas"
format: 
  revealjs:
    touch: false
    controls: true
    pointer:
      pointerSize: 32
    incremental: false
    slide-number: true
    css: styles.css
knitr:
  opts_chunk: 
    echo: true
    eval: true
    message: false
    warnings: false
    comment: "#>" 
revealjs-plugins:
  - pointer
filters: 
  - reveal-header
  - code-fullscreen
  - reveal-auto-agenda
---

# Installation

## Installing R

<https://cloud.r-project.org/>

## Installing RStudio

<https://posit.co/download/rstudio-desktop/>

## Installing RTools (Windows)

<https://cran.r-project.org/bin/windows/Rtools/>

## Posit Cloud

<https://posit.cloud/>

# R Packages

## Tidyverse

```{r}
install.packages("tidyverse")
```

## Plotly

```{r}
install.packages("plotly")
```

## ggthemes

```{r}
install.packages("ggthemes")
```

## Esquisse

```{r}
install.packages("esquisse")
```

## Load Packages

```{r}
library(tidyverse)
library(plotly)
library(ggthemes)
library(esquisse)
```

# R Projects

## R Projects

R Projects are designed to contain all your data, scripts, and output in one location. This ensures that things are compatible on other computers.

## Directories (Folders)

-   data

-   output

-   scripts

# RStudio Layout

## RStudio Layout

-   Scripts

-   Console

-   Environment

-   Files

-   Everything else

# Scripts

## Adding Script

-   Use the top right corner

# Loading Data

## Loading Data
